Qualitative Activity Recognition of Weight Lifting Exercises
============================================================

Synopsis
--------
Modern devises allow us to collect a large amount of data about personal activity relatively inexpensively. This project uses data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants to predict whether they are performing barbell lifts correctly or incorrectly in 5 different ways. 

Using the Random Forest algorithm, a model was created using 70% of the data for training and 30% for validation. The model produces a high out of sample accuracy of 99% when tested on the validation data set. 


Data Loading and Processing
---------------------------

The [training data set] (https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv) and the [test data set] (https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv) are downloaded from the course website.

We read both data sets.

```{r readData, cache=TRUE}
pmlTrain <- read.csv("pml-training.csv",na.strings=c("NA",""))
pmlTest <- read.csv("pml-testing.csv",na.strings=c("NA",""))

dim(pmlTrain)
dim(pmlTest)
```

Load the relevant libraries.

```{r}
library(caret)
library(randomForest)
```

Discard non-predictor columns -- x, all timestamps, user-name, new-window, num-window.

```{r}
rmCols <- grep("X|user_name|window|timestamp",names(pmlTrain))
pmlTrainSub <- pmlTrain[,-rmCols]
dim(pmlTrainSub)
```

Calculate proportion of NAs (nulls) for each column. 

```{r}
nRows <- nrow(pmlTrainSub)
NAcols <- apply(pmlTrainSub, 2, FUN= function(x) sum(is.na(x))/nRows)
```

Discard columns with high NAs (>= 50%).

```{r}
pmlTrainSub <- pmlTrainSub[,-which(NAcols >= 0.5)]
dim(pmlTrainSub)
```

We had trimmed the number of predictor columns to 52 (53 if including the class variable). The columns are:

```{r}
names(pmlTrainSub)
```



Train the Model
---------------

Create training set using 70% and validation set at 30% from training data to train the model.

```{r}
pmlTrainIndex <- createDataPartition(y = pmlTrainSub$classe, p=0.7,list=FALSE) 
pmlTrainData <- pmlTrainSub[pmlTrainIndex,]
pmlValData <- pmlTrainSub[-pmlTrainIndex,]

dim(pmlTrainData)
dim(pmlValData)
```

Fit a model using Random Forest. To minimize running time, we set the parameters to run using the "CV" method with 4 folds and also set the function to run in parallel mode. Run-time is approximately 8min.

```{r randomForest, cache=TRUE}
set.seed(13579)
modFit <- train(classe ~., data = pmlTrainData, method="rf", 
                trControl = trainControl(method = "cv", number = 4),
                allowParallel=T)
```

```{r}
modFit
```

Cross validate the prediction function using the validation data set to compute the relevant accuracy measures. 

```{r}
pred <- predict(modFit, newdata=pmlValData)
confusionMatrix(pred,pmlValData$classe)
```

The above results show that the out of sample accuracy is very high at **99%**, meaning that the expected out of sample error is **1%**.


Testing on New Data
-------------------

We now apply the model to the unseen test data to predict each record's class. 

```{r}
predTest <- predict(modFit, newdata=pmlTest)
```

The predictions are:

```{r}
predTest
```

Create a function to write each prediction into one file for submission to Coursera's Project Submission.

```{r}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
```

We now submit the prediction vector to the function to create the 20 out files in the working directory.

```{r}
pml_write_files(predTest)
```
